---
title: "Wrangling: strings"
sidebar: false
---

```{r message=FALSE, echo=FALSE}
library(tidyverse) # The stringr and forcats packages are part of tidyverse
```

# Learning goals

After this lesson, you should be able to:

- Manipulate and explore strings using the `stringr` package
- Construct regular expressions to find patterns in strings

<br>

To shape how we hold class today, go to my [PollEverywhere page](https://pollev.com/lesliemyint417) for some survey questions.

<br>

You can download a template Quarto file to start from [here](activity_templates/06-data-types-2.qmd). Save this template within the following directory structure:

- `your_course_folder`
    - `wrangling_data_types`
        - `code`
            - `05-data-types-1.qmd`
            - `06-data-types-2.qmd`


<br><br><br>


# Motivation: 30 Years of American Anxieties

In 2018 the data journalism organization [The Pudding](https://pudding.cool/) featured a story called [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) about themes in 30 years of posts to the Dear Abby column (an American advice column).

One way to understand themes in text data is to conduct a **qualitative analysis**, a methodology in which multiple readers read through instances of text several times to reach a consensus about themes.

Another way to understand themes in text data is to explore the text computationally with data science tools. This is what we will explore today. Both qualitative analysis and computational tools can be used in tandem. Often, using computational tools can help focus a close reading of select texts, which parallels the spirit of a qualitative analysis.

To prepare ourselves for a computational analysis, let's learn about strings.


# Strings

Strings are objects of the `character` class (abbreviated as `<chr>` in tibbles). When you print out strings, they display with double quotes:

```{r}
some_string <- "banana"
some_string
```

Working with strings generally will involve the use of **regular expressions**, a tool for finding patterns in strings. Regular expressions (regex, for short) look like the following:

```
"^the" (Strings that start with "the")
"end$" (Strings that end with "end")
```

Before getting to regular expressions, let's go over some fundamentals about working with strings. The `stringr` package (available within `tidyverse`) is great for working with strings.

## Creating strings

Creating strings by hand is useful for testing out regular expressions.

To create a string, type any text in either double quotes (`"`) or single quotes `'`. Using double or single quotes doesn't matter unless your string itself has single or double quotes.

```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

We can view these strings "naturally" (without the opening and closing quotes) with `str_view()`:

```{r}
str_view(string1)
str_view(string2)
```

**Exercise:** Create the string `It's Thursday`. What happens if you put the string inside single quotes? Double quotes?

```{r}
# Your code
```

<details>
  <summary>Solution</summary>

```{r error=TRUE}
x <- "It's Thursday" # We need double quotes because of the apostrophe
x
x <- 'It's Thursday'
```

</details>

Because `"` and `'` are special characters in the creation of strings, R offers another way to put them inside a string. We can **escape** these special characters by putting a `\` in front of them:

```{r}
string1 <- "This is a string with \"double quotes\""
string2 <- "This is a string with \'single quotes\'"
str_view(string1)
str_view(string2)
```

Given that `\` is a special character, how can we put the `\` character in strings? We have to escape it with `\\`.

**Exercise:** Create the string `C:\Users`. What happens when you don't escape the `\`?

```{r}
# Your code
```

<details>
  <summary>Solution</summary>

```{r error=TRUE}
x <- "C:\\Users"
str_view(x)

# \U is the start of special escape characters for Unicode characters
# The \U is expected to be followed by certain types of letters and numbers--like \U0928
x <- "C:\Users"
```

</details>

Other special characters include:

- `\t` (Creates a tab)
- `\n` (Creates a newline)

Both can be useful in plots to more neatly arrange text.

```{r}
string1 <- "Record temp:\t102"
string2 <- "Record temp:\n102"

str_view(string1)
str_view(string2)
```

**Exercise (Exploring function documentation):** Can we get `str_view()` to show the tab instead of `{\t}`? Enter `?str_view` in the Console to pull up the documentation for this function. Look through the arguments to see how we might do this.

<details>
  <summary>Solution</summary>

```{r eval=FALSE}
str_view(string1, html = TRUE)
```

</details>

**Reflection:** In your Process and Reflection Log, record any strategies that you learned about reading function documentation.

------------------

Often we will want to create new strings within data frames. We can use `str_c()` or `str_glue()`:

- With `str_c()` the strings to be combined are all separate arguments separated by commas.
- With `str_glue()` the desired string is written as a template with variable names inside curly braces `{}`.

```{r}
df <- tibble(
    first_name = c("Arya", "Olenna", "Tyrion", "Melisandre"),
    last_name = c("Stark", "Tyrell", "Lannister", NA)
)
df

df %>%
    mutate(
        full_name1 = str_c(first_name, " ", last_name),
        full_name2 = str_glue("{first_name} {last_name}")
    )
```

**Exercise:** In the following data frame, create a full date string in month-day-year format using both `str_c()` and `str_glue()`.

```{r}
df_dates <- tibble(
    year = c(2000, 2001, 2002),
    month = c("Jan", "Feb", "Mar"),
    day = c(3, 4, 5)
)
```

<details>
  <summary>Solution</summary>

```{r}
df_dates %>%
    mutate(
        date1 = str_c(month, "-", day, "-", year),
        date2 = str_glue("{month}-{day}-{year}")
    )
```

</details>

## Extracting information from strings

The `str_length()` counts the number of characters in a string.

```{r}
comments <- tibble(
    name = c("Alice", "Bob"),
    comment = c("The essay was well organized around the core message and had good transitions.", "Good job!")
)

comments %>%
    mutate(
        comment_length = str_length(comment)
    )
```

The `str_sub()` function gets a substring of a string. The 2nd and 3rd arguments indicate the beginning and ending position to extract.

- Negative positions indicate the position from the end of the word. (e.g., -3 indicates "3rd letter from the end")
- Specifying a position that goes beyond the word won't result in an error. `str_sub()` will just go as far as possible.

```{r}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
str_sub(x, -3, -1)
str_sub("a", 1, 5)
```

**Exercise:** Find the middle letter of each word in the data frame below. (Challenge: How would you handle words with an even number of letters?)

```{r}
df <- tibble(
    word_id = 1:3,
    word = c("replace", "match", "pattern")
)
```

<details>
  <summary>Solution</summary>

```{r}
df %>%
    mutate(
        word_length = str_length(word),
        middle_pos = ceiling(word_length/2),
        middle_letter = str_sub(word, middle_pos, middle_pos)
    )
```

</details>


## Finding patterns in strings with regular expressions

Suppose that you're exploring text data looking for places where people describe happiness. There are many ways to search. We could search for the word "happy" but that excludes "happiness" so we might search for "happi".

**Regular expressions** (regex) are a powerful language for describing patterns within strings.

```{r}
data(fruit)
data(words)
data(sentences)
```

We can use `str_view()` with the `pattern` argument to see what parts of a string match the regex supplied in the `pattern` argument. (Matches are enclosed in `<>`.)

```{r}
str_view(fruit, "berry")
```

**Essentials of forming a regex**

- Letters and numbers in a regex are matched exactly and are called **literal characters**.
- Most punctuation characters, like `.`, `+`, `*`, `[`, `]`, and `?`, have special meanings and are called **metacharacters**.
- **Quantifiers** come after a regex and control how many times a pattern can match:
    - `?`: match the preceding pattern 0 or 1 times
    - `+`: match the preceding pattern at least once
    - `*`: match the preceding pattern at least 0 times (any number of times)
    
**Exercise:** Before running the code below, predict what matches will be made. Run the code to check your guesses. Note that in all regex's below the `?, +, *` applies to the `b` only (not the `a`).

```{r}
str_view(c("a", "ab", "abb"), "ab?")
str_view(c("a", "ab", "abb"), "ab+")
str_view(c("a", "ab", "abb"), "ab*")
```

<details>
  <summary>Solution</summary>

```{r}
# This regex finds "a" then "b" at most once (can't have 2 or more b's in a row)
str_view(c("a", "ab", "abb"), "ab?")
```

```{r}
# There has to be an "a" followed by at least one b
# This is why the first string "a" isn't matched
str_view(c("a", "ab", "abb"), "ab+")
```

```{r}
# There must be an "a" and then any number of b's (including zero)
str_view(c("a", "ab", "abb"), "ab*")
```

</details>

- We can match any of a set of characters with `[]` (called a **character class**), e.g., `[abcd]` matches "a", "b", "c", or "d".
    - We can invert the match by starting with `^`: `[^abcd]` matches anything except "a", "b", "c", or "d".

```{r}
# Match words that have vowel-x-vowel
str_view(words, "[aeiou]x[aeiou]")

# Match words that have not_vowel-y-not_vowel
str_view(words, "[^aeiou]y[^aeiou]")
```

**Exercise** Using the `words` data, find words that have two vowels in a row followed by an "m".

```{r}
# Your code
```

<details>
  <summary>Solution</summary>

```{r}
str_view(words, "[aeiou][aeiou]m")
```

</details>

- The **alternation** operator `|` can be read just like the logical operator `|` ("OR") to pick between one or more alternative patterns. e.g., `apple|banana` searches for "apple" or "banana".

```{r}
str_view(fruit, "apple|melon|nut")
```

**Exercise:** Using the `fruit` data, find fruits that have a repeated vowel ("aa", "ee", "ii", "oo", or "uu".)

```{r}
# Your code
```

<details>
  <summary>Solution</summary>

```{r}
str_view(fruit, "aa|ee|ii|oo|uu")
```

</details>

- The `^` operator indicates the beginning of a string, and the `$` operator indicates the end of a string. e.g., `^a` matches strings that start with "a", and `a$` matches words that end with "a".
- Parentheses group together parts of a regular expression that should be taken as a bundle. (Much like parentheses in arithmetic statements.)
    - e.g., `ab+` is a little confusing. Does it match "ab" one or more times? Or does it match "a" first, then just "b" one or more times? (The latter, as we saw in an earlier example.) We can be very explicit and use `a(b)+`.
    
**Exercise:** Using the `words` data, find (1) words that start with "y" and (2) words that don't start with "y".

```{r}
# Your code
```

<details>
  <summary>Solution</summary>

```{r}
# Words that start with y
str_view(words, "^y")

# Words that don't start with y
str_view(words, "^[^y]")
```

</details>

------------------

The following are core `stringr` functions that use regular expressions:

- `str_view()` - View the first occurrence in a string that matches the regex
- `str_count()` - Count the number of times a regex matches within a string
- `str_detect()` - Determine if (TRUE/FALSE) the regex is found within string
- `str_subset()` - Return subset of strings that match the regex
- `str_extract(), str_extract_all()` - Return portion of each string that matches the regex. `str_extract()` extracts the first instance of the match. `str_extract_all()` extracts all matches.
- `str_replace(), str_replace_all()` - Replace portion of string that matches the regex with something else. `str_replace()` replaces the first instance of the match. `str_replace_all()` replaces all instances of the match.
- `str_remove(), str_remove_all()` - Removes the portion of the string that matches the pattern. Equivalent to `str_replace(x, "THE REGEX PATTERN", "")`

**Exercise:** Each person at your table should explore a different one of the functions (other than `str_view()`). Pull up the documentation page using `?function_name`. Explore the arguments and create a small example that demonstrates its usage. Share with your group members.


<br><br><br>


# Open-ended exploration: "Dear Abby" data

Read in the "Dear Abby" data underlying The Pudding's [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) article.

```{r}
posts <- read_csv("https://raw.githubusercontent.com/the-pudding/data/master/dearabby/raw_da_qs.csv")
```

Take a couple minutes to scroll throgh the [30 Years of American Anxieties](https://pudding.cool/2018/11/dearabby/) article to get ideas for themes that you might want to search for using regular expressions.

Using the `stringr` and regular expression tools that we talked about today, explore a theme of interest to you in the Dear Abby data.

**Reflection:** Make note of your exploration process. When do you tend to get stuck? How do you get unstuck? How do you notice "mistakes" or a need to change direction in the analysis? Record observations in your Process and Reflection Log.


<br><br><br>


# Project work time

I have one more poll question--navigate to my [PollEverywhere page](https://pollev.com/lesliemyint417).

Get together with your tentative project teammates from Tuesday.

1. If you are satisfied with your team composition, fill in your team details on this [Google Doc](https://docs.google.com/document/d/1BwgKdFXdWeaOiaDB-OGRcBhXM36qTIo79BmYZ9hvxko/edit?usp=sharing).
2. If you would like to continue searching for a team, you can do so today. Look through the Google Doc linked above as it's being filled out to get a sense for what team you'd like to join.

If you have already located a dataset relevant to one or more of your research questions and can read it into R, start to explore that data in working towards [Project Milestone 2](project.qmd#milestone-2).

Otherwise, peruse the [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) GitHub repository to find a dataset that is roughly (perhaps very roughly) related to your project domain. Start exploring this data in working towards [Project Milestone 2](project.qmd#milestone-2).

- **Why am I suggesting this?** Sometimes data that aren't perfectly aligned with our interests can still give useful context. Being able to provide meaning from "imperfect" data is a very useful skill because imperfect data are so common.



