---
title: "Wrangling: factors"
sidebar: false
---

# Learning goals

After this lesson, you should be able to:

- Recode and manage factors using the `forcats` package


<br><br><br>


# Factors

In R, factors are comprised of two components: the actual values of the data and the possible levels within the factor. Thus, to create a factor, you need to supply both these pieces of information.

For example, if we were to create a character vector of the twelve months, we could certainly do that:


```{r}
## Update this after doing ymd() processing on lakers
## all 12 months
all_months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

## our data
some_months <- c("Mar", "Dec", "Jan",  "Apr", "Jul")
```

However, if we were to sort this vector, R would sort this vector alphabetically.

```{r}
# alphabetical sort
sort(some_months)
```

While you and I know that this is not how months should be ordered, we haven’t yet told R that. To do so, we need to let R know that it’s a factor variable and what the levels of that factor variable should be.

```{r}
mon <- factor(some_months, levels = all_months)
mon
```

Here, we specify all the possible values that the factor could take in the levels = all_months argument. So, even though not all twelve months are included in the some_months object, we’ve stated that all of the months are possible values. Further, when you sort this variable, it now sorts in the sensical way!

**Update to use chickwts dataset?**

```{r eval=FALSE}
library(forcats)
fct_relevel()
fct_inorder() # skip this one?
fct_infreq()
fct_rev()
fct_reorder()
fct_recode()
ifelse() + factor()
```









