---
title: "Schedule"
sidebar: false
---

# Course Calendar

Readings in the schedule below refer to the following textbooks (freely available online):

- [R for Data Science (2e)](https://r4ds.hadley.nz/) by Wickham, Cetinkaya-Rundel, and Grolemund (Abbreviated as **R4DS**)
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/) (Abbreviated as **RPDS**)
- [Modern Data Science with R (3e)](https://mdsr-book.github.io/mdsr3e/) by Baumer, Kaplan, and Horton (Abbreviated as **MDSR**)

Guiding questions for the readings are available at the bottom of this page.

```{=html}
<table class="tg" data-quarto-disable-processing="true">
<thead>
  <tr>
    <th class="topleft" style="width: 10%"><b>Week</b></th>
    <th class="topleft" style="width: 30%"><b>Tuesday</b></th>
    <th class="topleft" style="width: 30%"><b>Thursday</b></th>
    <th class="topleft" style="width: 30%"><b>Announcements</b></th>
  </tr>
</thead>

<tbody>
  <tr>
    <td class="topleft">1</td>
    
    <td class="topleft">
      <b>9/5: Welcome! Meeting each other and designing our learning community</b> <br>
      Before class: Review the <a href="https://lmyint.github.io/212_fall_2023/syllabus.html">syllabus</a> and think about the questions posed in the green "Reflect" blocks.
    </td>
    <td class="topleft">
      <b>9/7: Advanced visualization in ggplot</b> <br>
      Before class: Review the construction of plots from STAT 112 and STAT 155. Answer the Guiding Questions at the bottom of this page. <br>
    </td>
    
    <td class="topleft">
      Work on <a href="https://lmyint.github.io/212_fall_2023/homework0.html">HW0</a> (your 10-year vision, doesn't need to be turned in).<br>Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a>
    </td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">2</td>
    <td class="topleft">
      <b>9/12: Advanced map visualization</b> <br>
      Before class: Watch <a href="https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx">this video</a> on Coordinate Reference Systems, and answer the Guiding Questions at the bottom of this page.<br>
    </td>
    
    <td class="topleft">
      <b>9/14: Advanced map visualization (continued)</b> <br>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a> by midnight on Wed 9/13. Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework2.html">HW2</a> due Wednesday 9/20 at midnight.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">3</td>
    <td class="topleft">
      <b>9/19: Interactive visualization</b> <br>
      Before class: Listen to <a href="https://datastori.es/127-storytelling-with-data-with-cole-nussbaumer-knaflic/">this podcast</a> from Chapter 7 (timestamp 18:09) through Chapter 8 (ending at timestamp 25:27). Answer the Guiding Question at the bottom of this page. Install the "shiny" and "plotly" R packages.
    </td>
    
    <td class="topleft">
      <b>9/21: Classroom Community and Connectedness (CC&C) Survey</b> <br>
      For the first 30 minutes, we will move our course projects forward. In the last hour of class, CC&C facilitators will come in to run an activity on how community-building is going in our course.
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework2.html">HW2</a> by midnight on Wednesday 9/20. Look ahead to HW3.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">4</td>
    <td class="topleft">
      <b>9/26: Data wrangling: numbers, logicals, and dates</b>
      Helpful readings (read before or after class): (All from R4DS) <a href="https://r4ds.hadley.nz/logicals.html">Chapter 13 (Logicals)</a>, <a href="https://r4ds.hadley.nz/numbers.html">Chapter 14 (Numbers)</a>, and <a href="https://r4ds.hadley.nz/datetimes.html">Chapter 18 (Dates/Times)</a>
    </td>
    
    <td class="topleft">
      <b>9/28: Data wrangling: strings</b> <br>
      Helpful readings (read before or after class): (All from R4DS) <a href="https://r4ds.hadley.nz/strings.html">Chapter 15 (Strings)</a> and <a href="https://r4ds.hadley.nz/regexps.html">Chapter 16 (Regular Expressions)</a>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework3.html">HW3</a>. Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework4.html">HW4</a> (Project Miletone 2).</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">5</td>
    <td class="topleft">
      <b>10/3: Data wrangling: factors</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/factors">Chapter 17 (Factors)</a> (R4DS).
    </td>
    
    <td class="topleft">
      <b>10/5: Writing functions</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/functions.html">R4DS Chapter 26 (Functions)</a> and <a href="https://bookdown.org/rdpeng/rprogdatascience/control-structures.html#if-else">RPDS Section 13.1 (if-else)</a>. 
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/project.html">Project Milestone 2</a> by either Wed 10/4 or Wed 10/11 at midnight. Start looking at <a href="https://lmyint.github.io/212_fall_2023/reflection1.html">Reflection 1</a>.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">6</td>
    <td class="topleft">
      <b>10/10: Loops and iteration</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/iteration.html">Chapter 27 (Iteration)</a> and <a href="https://www.rebeccabarter.com/blog/2019-08-19_purrr">this tutorial</a>. <br>
    </td>
    
    <td class="topleft">
      <b>10/12: Loops and iteration</b>
    </td>
    
    <td class="topleft">Turn in HW4 (Project Milestone 2) by Wed 10/11 if you haven't already. Turn in <a href="https://lmyint.github.io/212_fall_2023/reflection1.html">Reflection 1</a> by Wed 10/11.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">7</td>
    <td class="topleft">
      <b>10/17: Data acquisition: APIs</b> <br>
      Helpful readings (read before or after class):
    </td>
    
    <td class="topleft">
      <b>10/19: Data acquisition: Scraping</b> <br>
      Helpful readings (read before or after class): <a href="https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html">rvest vignette</a> <br>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework5.html">HW5</a>.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">8</td>
    <td class="topleft">
      <b>10/24: Project feedback</b> <br>
      <span class="project">Project progress presentation #1 (Milestone 3): Your team will present a 5-7 minute progress report and plan for next steps</span>
    </td>
    
    <td class="topleft">
      <b>10/26: No class - Fall Break 🍁</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">9</td>
    <td class="topleft">
      <b>10/31: Review and practice (Day 1)</b>
    </td>
    
    <td class="topleft">
      <b>11/2: Review and practice (Day 2)</b>
    </td>
    
    <td class="topleft">Work on <a href="https://lmyint.github.io/212_fall_2023/12-practice-review.html">review activities</a> as part of <a href="https://lmyint.github.io/212_fall_2023/homework6.html">HW6</a>.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">10</td>
    <td class="topleft">
      <b>11/7: Data acquisition: databases</b> <br>
      Required reading before class: <a href="https://r4ds.hadley.nz/databases.html">R4DS Chapter 22 (Databases)</a>.
    </td>
    
    <td class="topleft">
      <b>11/9: Data acquisition: databases (continued)</b>
    </td>
    
    <td class="topleft">Work on <a href="https://lmyint.github.io/212_fall_2023/homework7.html">HW7</a>.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">11</td>
    <td class="topleft">
      <b>11/14: Missing data: wrangling and missingness mechanisms</b> <br>
      Required reading before class: Chapters 1, 2, and 3 in <a href="https://tmb.njtierney.com/">The Missing Book</a>
    </td>
    
    <td class="topleft">
      <b>11/16: Missing data: imputation</b> <br>
      Required reading before class: Chapters 11, 13, and 14 in <a href="https://tmb.njtierney.com/">The Missing Book</a>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework7.html">HW7</a> on Wednesday 11/15.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">12</td>
    <td class="topleft">
      <b>11/21: Project presentations</b> <br>
      <span class="project">Project progress presentation #2 (Milestone 4): Your team will give a 10 minute presentation with intermediate results for 2 research questions.</span>
    </td>
    
    <td class="topleft">
      <b>11/23: No class - Thanksgiving Break 🦃</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">13</td>
    <td class="topleft">
      <b>11/28: Topics of your choosing from this point onwards (e.g., Python, GitHub, text analysis)</b>
    </td>
    
    <td class="topleft">
      <b>11/30: Topics of your choosing</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  
</tbody>
</table>
```


# Guiding Questions

Do your best to answer guiding questions before the indicated class period. Responses don't need to be turned in, but answering helps you prepare effectively for class.

## 11/7: Databases

Before class on Tuesday, install the `DBI` and `duckdb` packages. Post in the #questions channel on Slack if you run into issues.

As you read the [Databases](https://r4ds.hadley.nz/databases.html) chapter of R4DS, answer the following questions:

- What are the differences between client-server, cloud, and in-process database management systems (DBMSs)?
- Make note of how the following functions are used in a database workflow. What does each function do? What inputs/arguments does each function require? Create a set of notes that shows the sequence of these functions in a database workflow.
    - `DBI::dbConnect()`
    - `DBI::dbReadTable()`
    - `DBI::dbGetQuery()`
    - `tbl()`
    - `collect()`
- Make note of how the `showQuery()` function can help you learn SQL (structured query language) by translating `dplyr` code into SQL.
    - As you read Section 22.5 (SQL), create notes that relate parts of SQL queries to `dplyr` functions.


## 10/19: Web scraping

As you read the `rvest` package [vignette](https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html), answer the following questions:

- The first step of getting (scraping) data from an arbitrary web page is to read in the webpage. What `rvest` function(s) are relevant for this step?
- Next we need to select the HTML elements that contain the information that we want. What function(s) are relevant here?
    - How would we select all elements that have the class "author"?
    - How would we select all level 1 headers (tag `<h1>`)?
- Next we extract information from the selected HTML elements.
    - What is the difference between selecting the text contents of an HTML element and selecting an attribute of the HTML element? What functions are used for these two tasks?
    - How would we get all URLs for links that appear on a webpage?

## 9/19: Interactive visualization

After listening to [this podcast](https://datastori.es/127-storytelling-with-data-with-cole-nussbaumer-knaflic/) from Chapter 7 (timestamp 18:09) through Chapter 8 (ending at timestamp 25:27), reflect on the following question:

- What was new, unexpected, or interesting in the discussion about animations, interactivity, and dashboards?

<!-- Consider this post for another week: https://www.storytellingwithdata.com/blog/2014/04/exploratory-vs-explanatory-analysis -->

## 9/12: Advanced map visualization

After/while watching [this video](https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx) on Coordinate Reference Systems (CRS), answer the following questions:

- What is the shape of the Earth?
- Why is GDA94 a great datum name?
- What are the two components of a CRS/GCS?
- Why do we use many different local CRSs rather than just one CRS for the whole earth?
- Why is it insufficient to identify a location by its latitude and longitude?
- Why do we need to be mindful about CRSs when working with different spatial datasets?


## 9/7: Advanced visualization in ggplot

To review plot creation skills from STAT/COMP 112 and STAT 155, use the `diamonds` dataset in the `ggplot2` package to recreate the following visualizations:

```{r message=FALSE}
library(ggplot2)
data(diamonds)
```

```{r echo=FALSE, message=FALSE, fig.align="center", fig.width=12}
library(gridExtra)
p1 <- ggplot(diamonds, aes(x = cut)) + geom_bar() + theme_classic()
p2 <- ggplot(diamonds, aes(x = price)) + geom_histogram() + theme_classic()
p3 <- ggplot(diamonds, aes(x = price)) + geom_density() + theme_classic()
grid.arrange(p1, p2, p3, nrow = 1)

p4 <- ggplot(diamonds, aes(x = cut, y = price)) + geom_boxplot() + theme_classic()
p5 <- ggplot(diamonds, aes(x = price, color = cut)) + geom_density() + theme_classic()
grid.arrange(p4, p5, nrow = 1)

p6 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar() + theme_classic()
p7 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "dodge") + theme_classic()
p8 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "fill") + theme_classic()
grid.arrange(p6, p7, p8, nrow = 1)

p9 <- ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + theme_classic()
p10 <- ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_point() + geom_smooth(method = "lm") + theme_classic()
grid.arrange(p9, p10, nrow = 1)
```

