---
title: "Schedule"
sidebar: false
---

# Course Calendar

Readings in the schedule below refer to the following two textbooks (freely available online):

- [Modern Data Science with R (3e)](https://mdsr-book.github.io/mdsr3e/) by Baumer, Kaplan, and Horton (Abbreviated as **MDSR**)
- [R for Data Science (2e)](https://r4ds.hadley.nz/) by Wickham, Cetinkaya-Rundel, and Grolemund (Abbreviated as **R4DS**)

Guiding questions for the readings are available at the bottom of this page.

```{=html}
<table class="tg" data-quarto-disable-processing="true">
<thead>
  <tr>
    <th class="topleft" style="width: 10%"><b>Week</b></th>
    <th class="topleft" style="width: 30%"><b>Tuesday</b></th>
    <th class="topleft" style="width: 30%"><b>Thursday</b></th>
    <th class="topleft" style="width: 30%"><b>Announcements</b></th>
  </tr>
</thead>

<tbody>
  <tr>
    <td class="topleft">1</td>
    
    <td class="topleft">
      <b>9/5: Welcome! Meeting each other and designing our learning community</b> <br>
      Before class: Review the <a href="https://lmyint.github.io/212_fall_2023/syllabus.html">syllabus</a> and think about the questions posed in the green "Reflect" blocks.
    </td>
    <td class="topleft">
      <b>9/7: Advanced visualization in ggplot</b> <br>
      Before class: Review the construction of plots from STAT 112 and STAT 155. Answer the Guiding Questions at the bottom of this page. <br>
    </td>
    
    <td class="topleft">
      Work on <a href="https://lmyint.github.io/212_fall_2023/homework0.html">HW0</a> (your 10-year vision, doesn't need to be turned in).<br>Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a>
    </td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">2</td>
    <td class="topleft">
      <b>9/12: Advanced map visualization</b> <br>
      Before class: Watch <a href="https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx">this video</a> on Coordinate Reference Systems, and answer the Guiding Questions at the bottom of this page.<br>
    </td>
    
    <td class="topleft">
      <b>9/14: Interactive visualization (Part 1)</b> <br>
      Before class: TBD (NOT FINAL YET)
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a> by midnight on Wed 9/13. Look ahead to HW2.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">3</td>
    <td class="topleft">
      <b>9/19: Interactive visualization (Part 2)</b> <br>
      Before class: (NOT FINAL YET) Read up through the "Streamline computation" section in <a href="https://shiny.posit.co/r/getstarted/shiny-basics/lesson6/">this tutorial on reactive expressions</a> <br>
    </td>
    
    <td class="topleft">
      <b>9/21: Classroom Community and Connectedness Survey</b> <br>
      Before class: (NOT FINAL YET) Find a <a href="https://github.com/rfordatascience/tidytuesday">TidyTuesday dataset</a> that you would be interested in exploring for a mini-project. <br>
    </td>
    
    <td class="topleft">Turn in HW2. Look ahead to HW3.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">4</td>
    <td class="topleft">
      <b>9/26: Working with different data classes (Part 1)</b> <br>
      Before class: (NOT FINAL YET) Group 1 reads <a href="https://r4ds.hadley.nz/logicals.html">Chapter 13 (Logicals)</a> of R4DS. Group 2 reads <a href="https://r4ds.hadley.nz/numbers.html">Chapter 14 (Numbers)</a>. Group 3 reads <a href="https://r4ds.hadley.nz/datetimes.html">Chapter 18 (Dates/Times)</a>. <br>
    </td>
    
    <td class="topleft">
      <b>9/28: Working with different data classes (Part 2)</b> <br>
      Before class: (NOT FINAL YET) Read <a href="https://r4ds.hadley.nz/strings.html">Chapter 15 (Strings)</a>, <a href="https://r4ds.hadley.nz/regexps.html">Chapter 16 (Regular Expressions)</a>, and <a href="https://r4ds.hadley.nz/factors.html">Chapter 17 (Factors)</a> of R4DS. <br>
    </td>
    
    <td class="topleft">Turn in HW3. Look ahead to HW4.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">5</td>
    <td class="topleft">
      <b>10/3: Writing functions</b> <br>
      Before class: (NOT FINAL YET) Read <a href="https://r4ds.hadley.nz/functions.html">Chapter 26 (Functions)</a> and <a href="https://bookdown.org/rdpeng/rprogdatascience/control-structures.html#if-else">Section 13.1 (if-else)</a>. <br>
    </td>
    
    <td class="topleft">
      <b>10/5: Loops and iteration</b> <br>
      Before class: (NOT FINAL YET) Read <a href="https://r4ds.hadley.nz/iteration.html">Chapter 27 (Iteration)</a> and the following parts of <a href="https://www.rebeccabarter.com/blog/2019-08-19_purrr">this tutorial</a>: Up through (and including) "Simplest usage: repeated looping with map" <br>
    </td>
    
    <td class="topleft">Turn in HW4. Look ahead to HW5.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">6</td>
    <td class="topleft">
      <b>10/10: Iteration with purrr</b> <br>
      Before class: (NOT FINAL YET) Read the remainder of this <a href="https://www.rebeccabarter.com/blog/2019-08-19_purrr">purrr tutorial</a>. <br>
    </td>
    
    <td class="topleft">
      <b>10/12: Data acquisition: databases</b> <br>
      Before class: (NOT FINAL YET) Read <a href="https://r4ds.hadley.nz/databases.html">Chapter 22 (Databases)</a>. <br>
    </td>
    
    <td class="topleft">Turn in HW5. Look ahead to HW6.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">7</td>
    <td class="topleft">
      <b>10:17: Data acquisition: APIs</b> <br>
      Before class: (NOT FINAL YET) Read <a href="https://httr.r-lib.org/articles/quickstart.html">Getting started with httr</a> <br>
    </td>
    
    <td class="topleft">
      <b>10/19: Data acquisition: Scraping</b> <br>
      Before class: (NOT FINAL YET) Read the <a href="https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html">rvest vignette</a> <br>
    </td>
    
    <td class="topleft">Turn in HW6. Look ahead to HW7.</td>
  </tr>
</tbody>
</table>
```


# Guiding Questions

Do your best to answer guiding questions before the indicated class period. Responses don't need to be turned in, but answering helps you prepare effectively for class.

## 9/7: Advanced visualization in ggplot

To review plot creation skills from STAT/COMP 112 and STAT 155, use the `diamonds` dataset in the `ggplot2` package to recreate the following visualizations:

```{r message=FALSE}
library(ggplot2)
data(diamonds)
```

```{r echo=FALSE, message=FALSE, fig.align="center", fig.width=12}
library(gridExtra)
p1 <- ggplot(diamonds, aes(x = cut)) + geom_bar() + theme_classic()
p2 <- ggplot(diamonds, aes(x = price)) + geom_histogram() + theme_classic()
p3 <- ggplot(diamonds, aes(x = price)) + geom_density() + theme_classic()
grid.arrange(p1, p2, p3, nrow = 1)

p4 <- ggplot(diamonds, aes(x = cut, y = price)) + geom_boxplot() + theme_classic()
p5 <- ggplot(diamonds, aes(x = price, color = cut)) + geom_density() + theme_classic()
grid.arrange(p4, p5, nrow = 1)

p6 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar() + theme_classic()
p7 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "dodge") + theme_classic()
p8 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "fill") + theme_classic()
grid.arrange(p6, p7, p8, nrow = 1)

p9 <- ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + theme_classic()
p10 <- ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_point() + geom_smooth(method = "lm") + theme_classic()
grid.arrange(p9, p10, nrow = 1)
```

## 9/12: Advanced map visualization

After/while watching [this video](https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx) on Coordinate Reference Systems (CRS), answer the following questions:

- What is the shape of the Earth?
- Why is GDA94 a great datum name?
- What are the two components of a CRS/GCS?
- Why do we use many different local CRSs rather than just one CRS for the whole earth?
- Why is it insufficient to identify a location by its latitude and longitude?
- Why do we need to be mindful about CRSs when working with different spatial datasets?

## 9/14: Interactive visualization (Part 1)

After listening to [this podcast](https://datastori.es/127-storytelling-with-data-with-cole-nussbaumer-knaflic/) from Chapter 7 (timestamp 18:09) through Chapter 11 (ending at timestamp 33:41), reflect on the following questions:

- 

