---
title: "Schedule"
sidebar: false
---

# Course Calendar

Readings in the schedule below refer to the following textbooks (freely available online):

- [R for Data Science (2e)](https://r4ds.hadley.nz/) by Wickham, Cetinkaya-Rundel, and Grolemund (Abbreviated as **R4DS**)
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/) (Abbreviated as **RPDS**)
- [Modern Data Science with R (3e)](https://mdsr-book.github.io/mdsr3e/) by Baumer, Kaplan, and Horton (Abbreviated as **MDSR**)

Guiding questions for the readings are available at the bottom of this page.

```{=html}
<table class="tg" data-quarto-disable-processing="true">
<thead>
  <tr>
    <th class="topleft" style="width: 10%"><b>Week</b></th>
    <th class="topleft" style="width: 30%"><b>Tuesday</b></th>
    <th class="topleft" style="width: 30%"><b>Thursday</b></th>
    <th class="topleft" style="width: 30%"><b>Announcements</b></th>
  </tr>
</thead>

<tbody>
  <tr>
    <td class="topleft">1</td>
    
    <td class="topleft">
      <b>9/5: Welcome! Meeting each other and designing our learning community</b> <br>
      Before class: Review the <a href="https://lmyint.github.io/212_fall_2023/syllabus.html">syllabus</a> and think about the questions posed in the green "Reflect" blocks.
    </td>
    <td class="topleft">
      <b>9/7: Advanced visualization in ggplot</b> <br>
      Before class: Review the construction of plots from STAT 112 and STAT 155. Answer the Guiding Questions at the bottom of this page. <br>
    </td>
    
    <td class="topleft">
      Work on <a href="https://lmyint.github.io/212_fall_2023/homework0.html">HW0</a> (your 10-year vision, doesn't need to be turned in).<br>Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a>
    </td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">2</td>
    <td class="topleft">
      <b>9/12: Advanced map visualization</b> <br>
      Before class: Watch <a href="https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx">this video</a> on Coordinate Reference Systems, and answer the Guiding Questions at the bottom of this page.<br>
    </td>
    
    <td class="topleft">
      <b>9/14: Advanced map visualization (continued)</b> <br>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework1.html">HW1</a> by midnight on Wed 9/13. Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework2.html">HW2</a> due Wednesday 9/20 at midnight.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">3</td>
    <td class="topleft">
      <b>9/19: Interactive visualization</b> <br>
      Before class: Listen to <a href="https://datastori.es/127-storytelling-with-data-with-cole-nussbaumer-knaflic/">this podcast</a> from Chapter 7 (timestamp 18:09) through Chapter 8 (ending at timestamp 25:27). Answer the Guiding Question at the bottom of this page. Install the "shiny" and "plotly" R packages.
    </td>
    
    <td class="topleft">
      <b>9/21: Classroom Community and Connectedness (CC&C) Survey</b> <br>
      For the first 30 minutes, we will move our course projects forward. In the last hour of class, CC&C facilitators will come in to run an activity on how community-building is going in our course.
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework2.html">HW2</a> by midnight on Wednesday 9/20. Look ahead to HW3.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">4</td>
    <td class="topleft">
      <b>9/26: Data wrangling: numbers, logicals, and dates</b>
      Helpful readings (read before or after class): (All from R4DS) <a href="https://r4ds.hadley.nz/logicals.html">Chapter 13 (Logicals)</a>, <a href="https://r4ds.hadley.nz/numbers.html">Chapter 14 (Numbers)</a>, and <a href="https://r4ds.hadley.nz/datetimes.html">Chapter 18 (Dates/Times)</a>
    </td>
    
    <td class="topleft">
      <b>9/28: Data wrangling: strings</b> <br>
      Helpful readings (read before or after class): (All from R4DS) <a href="https://r4ds.hadley.nz/strings.html">Chapter 15 (Strings)</a> and <a href="https://r4ds.hadley.nz/regexps.html">Chapter 16 (Regular Expressions)</a>
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/homework3.html">HW3</a>. Look ahead to <a href="https://lmyint.github.io/212_fall_2023/homework4.html">HW4</a> (Project Miletone 2).</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">5</td>
    <td class="topleft">
      <b>10/3: Data wrangling: factors</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/factors">Chapter 17 (Factors)</a> (R4DS).
    </td>
    
    <td class="topleft">
      <b>10/5: Writing functions</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/functions.html">R4DS Chapter 26 (Functions)</a> and <a href="https://bookdown.org/rdpeng/rprogdatascience/control-structures.html#if-else">RPDS Section 13.1 (if-else)</a>. 
    </td>
    
    <td class="topleft">Turn in <a href="https://lmyint.github.io/212_fall_2023/project.html">Project Milestone 2</a> by either Wed 10/4 or Wed 10/11 at midnight. Start looking at <a href="https://lmyint.github.io/212_fall_2023/reflection1.html">Reflection 1</a>.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">6</td>
    <td class="topleft">
      <b>10/10: Loops and iteration</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/iteration.html">Chapter 27 (Iteration)</a> and <a href="https://www.rebeccabarter.com/blog/2019-08-19_purrr">this tutorial</a>. <br>
    </td>
    
    <td class="topleft">
      <b>10/12: Loops and iteration</b>
    </td>
    
    <td class="topleft">Turn in HW4 (Project Milestone 2) by Wed 10/11 if you haven't already. Turn in <a href="https://lmyint.github.io/212_fall_2023/reflection1.html">Reflection 1</a> by Wed 10/11.</td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">7</td>
    <td class="topleft">
      <b>10/17: Data acquisition: APIs</b> <br>
      Helpful readings (read before or after class):
    </td>
    
    <td class="topleft">
      <b>10/19: Data acquisition: Scraping</b> <br>
      Helpful readings (read before or after class): <a href="https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html">rvest vignette</a> <br>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">8</td>
    <td class="topleft">
      <b>10/24: Project feedback & styling your personal site with HTML and CSS</b> <br>
      <span class="project">Project progress presentation #1: Your team will present a 5-7 minute progress report and plan for next steps</span> <br>
      After project presentations, we will learn a bit more about HTML and CSS so that you can style your personal websites.
    </td>
    
    <td class="topleft">
      <b>10/26: No class - Fall Break 🍁</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">9</td>
    <td class="topleft">
      <b>10/31: Data acquisition: databases</b> <br>
      Helpful readings (read before or after class): <a href="https://r4ds.hadley.nz/databases.html">R4DS Chapter 22 (Databases)</a>.
    </td>
    
    <td class="topleft">
      <b>11/2: Causal inference and machine learning in data science</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">10</td>
    <td class="topleft">
      <b>11/7: Missing data: wrangling and missingness mechanisms</b> <br>
      Helpful readings (read before or after class): 
    </td>
    
    <td class="topleft">
      <b>11/9: Missing data: imputation</b>
    </td>
    
    <td class="topleft"></td>
  </tr>
  <!----------------------------------------------------------->
  
  
  <tr>
    <td class="topleft">11</td>
    <td class="topleft">
      <b>11/14: Likely Python, GitHub, and topics of your choosing from this poitn onward</b> <br>
    </td>
    
    <td class="topleft">
      <b>11/16: Project presentations</b> <br>
      <span class="project">Project progress presentation #2: Your team will give a 10 minute presentation with intermediate results for 2 research questions.</span>
    </td>
    
    <td class="topleft"></td>
  </tr>
  
</tbody>
</table>
```


# Guiding Questions

Do your best to answer guiding questions before the indicated class period. Responses don't need to be turned in, but answering helps you prepare effectively for class.

## 9/7: Advanced visualization in ggplot

To review plot creation skills from STAT/COMP 112 and STAT 155, use the `diamonds` dataset in the `ggplot2` package to recreate the following visualizations:

```{r message=FALSE}
library(ggplot2)
data(diamonds)
```

```{r echo=FALSE, message=FALSE, fig.align="center", fig.width=12}
library(gridExtra)
p1 <- ggplot(diamonds, aes(x = cut)) + geom_bar() + theme_classic()
p2 <- ggplot(diamonds, aes(x = price)) + geom_histogram() + theme_classic()
p3 <- ggplot(diamonds, aes(x = price)) + geom_density() + theme_classic()
grid.arrange(p1, p2, p3, nrow = 1)

p4 <- ggplot(diamonds, aes(x = cut, y = price)) + geom_boxplot() + theme_classic()
p5 <- ggplot(diamonds, aes(x = price, color = cut)) + geom_density() + theme_classic()
grid.arrange(p4, p5, nrow = 1)

p6 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar() + theme_classic()
p7 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "dodge") + theme_classic()
p8 <- ggplot(diamonds, aes(x = cut, fill = color)) + geom_bar(position = "fill") + theme_classic()
grid.arrange(p6, p7, p8, nrow = 1)

p9 <- ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + theme_classic()
p10 <- ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_point() + geom_smooth(method = "lm") + theme_classic()
grid.arrange(p9, p10, nrow = 1)
```

## 9/12: Advanced map visualization

After/while watching [this video](https://youtu.be/xJyJlKbZFlc?si=pc7ThQPO-v8ur-qx) on Coordinate Reference Systems (CRS), answer the following questions:

- What is the shape of the Earth?
- Why is GDA94 a great datum name?
- What are the two components of a CRS/GCS?
- Why do we use many different local CRSs rather than just one CRS for the whole earth?
- Why is it insufficient to identify a location by its latitude and longitude?
- Why do we need to be mindful about CRSs when working with different spatial datasets?

## 9/19: Interactive visualization

After listening to [this podcast](https://datastori.es/127-storytelling-with-data-with-cole-nussbaumer-knaflic/) from Chapter 7 (timestamp 18:09) through Chapter 8 (ending at timestamp 25:27), reflect on the following question:

- What was new, unexpected, or interesting in the discussion about animations, interactivity, and dashboards?

<!-- Consider this post for another week: https://www.storytellingwithdata.com/blog/2014/04/exploratory-vs-explanatory-analysis -->

